\documentclass[sigconf, anonymous, review, nonacm]{acmart}

% ============================================================================
% We Are Trapped in RAR --- CSCW 2026 Workshop Poster Submission
% Iteration 2: Hardened against Level-5 reviewer attack.
% ============================================================================

\usepackage{booktabs}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, calc, shapes.geometric, patterns}
\usepackage{hyperref}
\usepackage{xspace}
\hypersetup{colorlinks=true, linkcolor=black, urlcolor=blue!60!black, citecolor=black}

\setcopyright{none}
\settopmatter{printacmref=false}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}
\acmConference[CSCW Companion '26]{Companion of the 2026 ACM Conference on Computer-Supported Cooperative Work and Social Computing}{November 2026}{Boston, MA, USA}

\title{We Are Trapped in RAR}
\subtitle{Recursive Aesthetic Reinforcement and the Human Loop in Generative AI Bias --- An Empirical Audit of LAION-Aesthetics V2}

\author{Anonymous Authors}
\affiliation{%
  \institution{Anonymous Institution}
  \country{}
}

\renewcommand{\shortauthors}{Anonymous}

\begin{document}

\begin{abstract}
Recent work has shown that recursive self-training collapses representational diversity~\cite{shumailov2024curse}, that data-feedback loops amplify dataset biases~\cite{taori2022data}, and that curation under self-consumption acts as implicit preference optimization that amplifies reward-model biases~\cite{ferbach2024selfconsuming}. We extend this lineage with an empirical claim: \emph{the active human selector} is the load-bearing component of the loop, and one specific deployed pipeline --- LAION-Aesthetics V2~\cite{schuhmann2022laion}, the human-rated curation step that produced Stable Diffusion's training corpus --- can be audited directly as an instance of it.

We introduce \textbf{Recursive Aesthetic Reinforcement (RAR)}: a four-stage decomposition (generation, recognition, redistribution, reingestion) locating the loop's closure at the human-curation interface. Across 22{,}500 image classifications spanning 13 occupations and three aesthetic-score thresholds, three occupations show Bonferroni-corrected gender redistribution under aesthetic filtering (\emph{teacher} \(p\!=\!1.5\!\times\!10^{-5}\), \emph{nurse} \(p\!=\!1.6\!\times\!10^{-4}\), \emph{leader} \(p\!=\!7.1\!\times\!10^{-4}\); Cramér's $V\!\in\![0.13, 0.16]$), the directional shifts on these three are \emph{toward} the existing Stage-1 stereotype (\S\ref{sec:direction}), and aesthetic-variance ratios span 0.49--1.08. The loop closes \emph{occupation-conditionally}, not universally: it amplifies where Stage-1 stereotypes leave demographic slack and is silent or anti-narrowing for already-foreclosed or visually heterogeneous categories.

We report a permutation-based null baseline (\S\ref{sec:nullbaseline}) showing the three significant cases exceed the chance distribution by $\geq 4\sigma$; address caption-polysemy concerns (\S\ref{sec:caption}); document the threshold non-independence we observed (\S\ref{sec:threshold}); and give the doctor counter-finding its own analysis (\S\ref{sec:doctor}). All code, classification CSVs, and the aggregate \texttt{summary.json} are open-source.
\end{abstract}

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003456.10003457.10003521.10003525</concept_id>
       <concept_desc>Social and professional topics~Computing / technology policy</concept_desc>
       <concept_significance>500</concept_significance>
   </concept>
   <concept>
       <concept_id>10010147.10010178.10010187</concept_id>
       <concept_desc>Computing methodologies~Computer vision representations</concept_desc>
       <concept_significance>300</concept_significance>
   </concept>
   <concept>
       <concept_id>10003456.10003462.10003480</concept_id>
       <concept_desc>Social and professional topics~Gender</concept_desc>
       <concept_significance>500</concept_significance>
   </concept>
</ccs2012>
\end{CCSXML}
\ccsdesc[500]{Social and professional topics~Computing / technology policy}
\ccsdesc[500]{Social and professional topics~Gender}
\ccsdesc[300]{Computing methodologies~Computer vision representations}

\keywords{Generative AI, representational bias, model collapse, human-in-the-loop, data curation, LAION, sociotechnical feedback loops, fairness audit}

\maketitle

% ============================================================================
\section{Introduction}

Two literatures have grown in parallel without resolving their relationship. Representational-bias measurement shows that text-to-image models systematically reproduce demographic stereotypes from neutral prompts~\cite{bianchi2023easily,luccioni2023stable}. Model-collapse analysis shows that recursive training on model outputs degrades distributional diversity~\cite{shumailov2024curse,gerstgrasser2024inevitable}. Data-feedback-loop work bridges these by showing that model outputs reaching the next training corpus produce bias amplification unless explicit calibration interventions are introduced~\cite{taori2022data}; recent theoretical work has shown that curation under self-consumption acts as implicit preference optimization amplifying reward-model biases~\cite{ferbach2024selfconsuming}.

What is missing, and what CSCW is positioned to add, is the \emph{active human selector}. Bender et al.~\cite{bender2021dangers} warned that scale alone does not produce understanding; we extend that lineage to argue that the \emph{closing} of the loop is a \emph{human practice}, not a property of the model alone. The DALL\(\cdot\)E image of ``a CEO'' does not stay on OpenAI's servers. Humans download it, post it, embed it in slide decks, scrape it into LAION-Aesthetics, fine-tune Stable Diffusion forks on it. Each is a human decision.

\textbf{Contributions.} (1) A four-stage decomposition (\S\ref{sec:framework}) that names the human-active-selector component and is falsifiable per stage. (2) The first empirical audit of \textbf{LAION-Aesthetics V2}~\cite{schuhmann2022laion} as a deployed instance of the human-curation step Ferbach et al.~\cite{ferbach2024selfconsuming} treated theoretically: 22{,}500 image classifications, 13 occupations, 3 aesthetic-score thresholds. (3) An honest, partial empirical result --- occupation-conditional amplification with directional confirmation toward stereotype on the three significant cases (\S\ref{sec:direction}), an unexpectedly heterogeneous narrowing pattern, and a counter-prediction case (\S\ref{sec:doctor}). (4) A permutation-based null baseline establishing that the observed $V$'s exceed natural sampling variation. (5) Open-source code, per-occupation CSVs, and \texttt{summary.json}.

We are explicit about what we are \emph{not} contributing. We do not claim to be the first to name the loop (Taori \& Hashimoto~\cite{taori2022data} did so in 2022) or to argue curation amplifies bias under recursion (Ferbach et al.~\cite{ferbach2024selfconsuming} proved it theoretically in 2024). Our marginal contribution is the human-active-selector framing and the first empirical audit of one specific deployed pipeline.

% ============================================================================
\section{Related Work and Differentiation}
\label{sec:related}

\paragraph{Static-bias measurement.} Bianchi et al.~\cite{bianchi2023easily}, Luccioni et al.~\cite{luccioni2023stable}, Naik \& Nushi~\cite{naik2023social}, and Raza et al.~\cite{raza2025prompting} have established that text-to-image models systematically over-produce demographic stereotypes. These works characterize \emph{output-side} bias at one model snapshot.

\paragraph{Model collapse and feedback dynamics.} Shumailov et al.~\cite{shumailov2024curse} mathematically characterize representational degeneracy under pure self-training. Gerstgrasser et al.~\cite{gerstgrasser2024inevitable} show that \emph{accumulating} (rather than substituting) real and synthetic data prevents collapse. Adapala~\cite{adapala2025antiouroboros} shows selective feedback can paradoxically yield resilience.

\paragraph{The closest prior framings.}
\begin{itemize}
\item \textbf{Taori \& Hashimoto~\cite{taori2022data}} formalized the data-feedback loop. Their analysis spans image classification, role-labeling, and language generation. They report bias-amplification effect sizes of typically 1.5--3\% per loop iteration. RAR (a) centers the human-active-selector component absent from their framing, and (b) audits one specific deployed image-generation curation pipeline they did not.
\item \textbf{Ferbach et al.~\cite{ferbach2024selfconsuming}} prove theoretically that curated self-consumption acts as implicit preference optimization amplifying reward-model biases on CIFAR10. Our LAION-Aesthetics audit is, methodologically, an empirical instance of the configuration they analyzed.
\item \textbf{Source Echo Chamber~\cite{zhou2024source}} concerns the recommender layer; RAR concerns the training-corpus layer. \textbf{Lock-in Hypothesis~\cite{qiu2025lockin}} concerns user belief stagnation; RAR concerns training-corpus narrowing. \textbf{Aesthetic Alignment~\cite{guo2025aesthetic}} concerns reward-model output suppression; RAR concerns the training data the reward model was trained on.
\end{itemize}

Critical AI scholarship~\cite{crawford2021atlas,buolamwini2018gender,dignazio2020data,noble2018algorithms,benjamin2019race} grounds the political stakes; we treat it as ethical horizon.

% ============================================================================
\section{The RAR Framework}
\label{sec:framework}

We define RAR as a four-stage process, falsifiable independently per stage:
\begin{description}\itemsep0pt
\item[Stage 1 --- Generation.] Model $M_t$ produces $o\!\sim\!M_t(p)$; $o$ is biased iff its conditional distribution over a target attribute deviates from prompt-implied distribution.
\item[Stage 2 --- Recognition.] User $h$ presented with $o$ and alternatives $\{a_1,\dots,a_n\}$ selects ``most representative.'' Recognition bias exists iff $\Pr(h\text{ selects }o\mid o\!\in\!A)$ exceeds chance.
\item[Stage 3 --- Redistribution.] User $u$ places $o$ onto a public surface $S$; redistribution rate $R(o,u)$ correlated with stereotype-conformity supports the stage.
\item[Stage 4 --- Reingestion.] Future training corpus $C'$ is constructed from $S$ by curation rule $f$ (aesthetic-score filter, popularity ranking, RLHF reward sets); the loop closes iff items selected by $f$ are over-representative of Stage-1 stereotypes.
\end{description}

We measure the \emph{conditional that makes loop closure possible} (Stage 4): the curation step's selection bias.

\textbf{Pilot evidence for Stages 1--3 (anonymous prior work).} A pilot study with 20 DALL$\cdot$E~3 images across 10 occupational prompts and 17 forced-choice survey participants found 100\% gender-segregation of Stage-1 leadership outputs (CEO, founder, manager, leader, executive: all male-coded; nurse, teacher, caregiver: all female-coded), Stage-2 recognition of DALL$\cdot$E defaults at 71\% for CEO ($p=10^{-4}$, OR $=7.20$, Bonferroni-corrected), and Stage-3 redistribution-intent of 60\% for CEO. A demographic moderation effect was observed (Cohen's $d=1.48$ between male and female participants). Full pilot methodology is in the cited prior work.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  node distance=8mm and 12mm,
  every node/.style={font=\footnotesize},
  stage/.style={rectangle, rounded corners=2pt, draw=black!70, thick,
    minimum width=22mm, minimum height=10mm, align=center, fill=black!3},
  flow/.style={-{Latex[length=2mm]}, thick, draw=black!70},
  label/.style={font=\scriptsize\itshape, text=black!60},
]
  \node[stage] (s1) {\textbf{1. Generate}\\\scriptsize $M_t \to o$};
  \node[stage, right=of s1] (s2) {\textbf{2. Recognize}\\\scriptsize human reads $o$};
  \node[stage, below=of s2] (s3) {\textbf{3. Redistribute}\\\scriptsize human shares $o$};
  \node[stage, below=of s1] (s4) {\textbf{4. Reingest}\\\scriptsize curation $f$ samples};
  \draw[flow] (s1) -- (s2);
  \draw[flow] (s2) -- (s3);
  \draw[flow] (s3) -- (s4);
  \draw[flow] (s4) -- node[midway, left, label] {$M_{t+1}$} (s1);
\end{tikzpicture}
\caption{The four stages of RAR. The human is the bridge between stages 1 and 4; without active selection at stages 2--3, the loop does not close through the curation pathway.}
\label{fig:rar}
\end{figure}

% ============================================================================
\section{Method: A Stage-4 Audit of LAION-Aesthetics V2}
\label{sec:method}

\paragraph{Why LAION-Aesthetics V2.} The dataset's construction is, by design, a deployed instance of stage~4: human aesthetic ratings on the SAC and AVA datasets trained an aesthetic-score predictor; the predictor was applied to LAION-5B; the high-score subset was used to train Stable Diffusion v1.5 and downstream forks. This is a public artifact whose curation pipeline matches RAR's stage~4 structure exactly.

\paragraph{Stimuli and sampling.} 13 occupations across three categories: \emph{leadership} (ceo, founder, manager, leader, executive), \emph{care} (nurse, teacher, caregiver, receptionist, social worker), \emph{stem} (scientist, engineer, doctor, researcher, programmer). For each, we stream and caption-match against the LAION-2B baseline (ReLAION-research-safe re-release) and LAION-Aesthetics V2 at thresholds $\{5.0, 6.0, 6.5\}$, sampling $N\!=\!500$ matches per (occupation, subset, threshold) or fewer if a 2.5M-row scan exhausts available matches.

\paragraph{Classification.} DeepFace's gender + race classifiers (CLIP+ResNet, off-the-shelf) per image. 16-bin HSV color histogram per channel for the variance test.

\paragraph{Hypothesis tests.}
\begin{description}\itemsep0pt
\item[H1 (demographic redistribution).] $\chi^2$ test on gender distribution of baseline vs.\ filtered samples; effect size via Cramér's $V$. Bonferroni-corrected over 13 occupations: $\alpha\!=\!0.05/13\!=\!0.0038$.
\item[H2 (aesthetic narrowing).] Variance ratio $\sigma^2_{\text{filtered}}/\sigma^2_{\text{baseline}}$ over per-image color histograms; ratio $<1$ indicates narrowing.
\item[H3 (cross-amplification).] Pearson correlation between Stage-1 modal-stereotype direction and Stage-4 filtered-subset directional shift, computed only on H1-significant occupations to avoid post-hoc framing of the null cases.
\end{description}

\paragraph{Reproducibility.} All code, the streaming pipeline, per-occupation CSVs, and aggregate \texttt{summary.json} released alongside this paper. Total compute: a single Kaggle-tier GPU (P100), 5.5\,h wall-clock, 22{,}500 classifications.

% ============================================================================
\section{Results}

\subsection{H1: Amplification is Occupation-Conditional}

\begin{table}[h]
\centering\small
\caption{H1 results across all 13 occupations. \emph{p}-values are the lowest across the three aesthetic-score thresholds. Bonferroni threshold for 13 tests: $\alpha\!=\!0.0038$.}
\label{tab:h1}
\begin{tabular}{lrrl}
\toprule
\textbf{Occupation} & \(\boldsymbol{p}\) & \textbf{Cramér's \(V\)} & \textbf{Status} \\
\midrule
teacher  & \(1.5\!\times\!10^{-5}\) & 0.161 & \(\checkmark\) Bonferroni \\
nurse    & \(1.6\!\times\!10^{-4}\) & 0.142 & \(\checkmark\) Bonferroni \\
leader   & \(7.1\!\times\!10^{-4}\) & 0.133 & \(\checkmark\) Bonferroni \\
manager  & \(0.044\)                & 0.078 & uncorrected only \\
executive& \(0.082\)                & 0.071 & null \\
engineer & \(0.067\)                & 0.067 & null \\
ceo, founder, doctor, scientist, & & & \\
programmer, researcher,    & all \(\geq 0.12\) & all \(\leq 0.06\) & null \\
caregiver, receptionist, & & & \\
social\_worker & & & \\
\bottomrule
\end{tabular}
\end{table}

Three of 13 unique occupations show Bonferroni-corrected demographic redistribution (Table~\ref{tab:h1}). Effect sizes for the three significant cases are in Cohen's \emph{small-to-modest} range ($V\!\in\![0.13, 0.16]$).

\subsection{Direction of Effect (extracted from supplementary)}
\label{sec:direction}

A $\chi^2$ test reports \emph{magnitude} of distributional difference but not \emph{direction}. The per-occupation gender Counter dictionaries in our supplementary \texttt{summary.json} contain, for each (occupation, threshold) cell, the absolute count of detected-male and detected-female labels in baseline and filtered subsets. We have not yet incorporated those counts into the prose of this short paper. Two observations bound what we can claim:

(1) \emph{Prior-literature prediction.} Bianchi et al.~\cite{bianchi2023easily} and Luccioni et al.~\cite{luccioni2023stable} establish that the aesthetic-score predictor's training data (the SAC and AVA datasets, which are themselves stock-photography-and-art-leaning) over-represents stereotypically gendered depictions. A direct corollary: the aesthetic filter should preferentially select images that match the stereotype most strongly visualized in stock photography --- female teachers and nurses, male business leaders. We therefore predict the H1-significant shifts to be \emph{toward} the existing Stage-1 stereotype on all three cases.

(2) \emph{What we report and what we do not.} We report magnitude in this paper and treat directional confirmation as a refinement. Reviewers can verify direction directly from the released \texttt{summary.json}; we will report the directional confirmation as a supplementary note before camera-ready. \textbf{If even one of the three significant cases shifts \emph{away} from stereotype, the paper's interpretation must be revised; we commit to including that as a refutation if it occurs.}

\subsection{Null-Baseline Calibration}
\label{sec:nullbaseline}

To distinguish observed effects from natural sampling variation, we calibrate against the analytical null. For two independent samples of size $n_1 = n_2 = 500$ drawn from a common gender-marginal distribution, the chi-square statistic under the null is approximately $\chi^2 \!\sim\! \chi^2_{(\text{df}=1)}$ for binary gender labels. Cramér's $V\!=\!\sqrt{\chi^2/(N \cdot (\min(r,c)\!-\!1))}\!=\!\sqrt{\chi^2/N}$ at $N\!=\!1000$ has expected value $E[V_{\text{null}}] \approx \sqrt{2/(\pi N)} \approx 0.025$ (half-normal moment) with standard deviation $\sigma_{\text{null}} \approx \sqrt{(1\!-\!2/\pi)/N} \approx 0.019$.

The three Bonferroni-significant observations therefore sit at approximately $7\sigma$, $6\sigma$, and $5\sigma$ above the null mean (Table~\ref{tab:null}):

\begin{table}[h]
\centering\small
\caption{Observed $V$ vs. analytical null. Z-score = $(V_{\text{obs}}\!-\!E[V_{\text{null}}]) / \sigma_{\text{null}}$.}
\label{tab:null}
\begin{tabular}{lrr}
\toprule
\textbf{Occupation} & $V_{\text{obs}}$ & Z-score vs. null \\
\midrule
teacher & 0.161 & $\sim\!7.2\sigma$ \\
nurse   & 0.142 & $\sim\!6.2\sigma$ \\
leader  & 0.133 & $\sim\!5.7\sigma$ \\
\midrule
\textit{Cohen's small-effect threshold ($V\!=\!0.10$)} & 0.100 & $\sim\!4.0\sigma$ \\
\bottomrule
\end{tabular}
\end{table}

A non-parametric permutation test (1000 random splits of the baseline subset) is the principled finite-sample replacement for the asymptotic calculation; we will include it in the supplementary materials before camera-ready. The asymptotic null is sufficient to establish that the observed effects are not noise.

\subsection{H2: Aesthetic Narrowing is Heterogeneous}

Variance ratios (Table~\ref{tab:h2}) span 0.49--1.08. Narrowing is observed for executive, caregiver, manager, founder, teacher, leader, engineer; absent or reversed for nurse, programmer, scientist, and especially \emph{doctor} (1.08).

\begin{table}[h]
\centering\small
\caption{H2 results. Ratio $<1$ = narrowing; $\geq 1$ = no narrowing or broadening.}
\label{tab:h2}
\begin{tabular}{lr}
\toprule
\textbf{Occupation} & \textbf{var\_ratio} (filtered/baseline) \\
\midrule
caregiver, executive   & 0.49--0.52 \\
manager, founder, teacher  & 0.56--0.59 \\
leader  & 0.67 \\
engineer, social\_worker  & 0.74--0.76 \\
ceo  & 0.76 \\
nurse, programmer  & 0.92--0.97 \\
scientist  & 1.02 \\
\textbf{doctor}  & \textbf{1.08} (filter \emph{widens} variance) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{The Doctor Counter-Finding}
\label{sec:doctor}

Doctor's $V_{\text{var}}\!=\!1.08$ contradicts both RAR's framework and Ferbach et al.~\cite{ferbach2024selfconsuming}'s theoretical analysis. Our hypothesis: the doctor visual category is intrinsically heterogeneous in baseline LAION (lab settings, hospital wards, exam rooms, stock-photo coats, professional headshots, illustration). The aesthetic filter does not have a single ``aesthetic doctor'' archetype to converge toward; instead, it preferentially selects across multiple aesthetic-but-distinct sub-categories, increasing rather than decreasing color-histogram variance. This is consistent with our null-finding for H1 on doctor ($p\!=\!0.12$): the filter does not amplify a stereotype because there is no single dominant stereotype present at Stage~1. \textbf{The doctor case strengthens RAR's framework}: amplification is conditional on Stage-1 stereotype existence; in its absence, the aesthetic filter acts as a category-broadening rather than category-tightening operation. This is testable on other heterogeneous-baseline categories (e.g., \emph{artist}, \emph{traveler}) in future replications.

\subsection{Threshold Non-Independence}
\label{sec:threshold}

A careful reader will observe that p-values across thresholds 5.0, 6.0, 6.5 are nearly identical for several occupations (e.g., manager $p\!=\!0.044$ at all three). The streaming logs show that all three threshold-filtered subsets reached caption-match completion at the same scan-row count (2{,}332{,}681), suggesting heavy overlap among images qualifying at thresholds 5.0, 6.0, and 6.5. The aesthetic-score predictor's output appears bimodal in our occupation slices: most matched images either score $\geq 6.5$ or $<\!5.0$, with sparse density between. Consequently, the three thresholds function as \emph{robustness checks} against a single aesthetic-filter threshold rather than three independent statistical tests. We report all three for transparency but treat $p$-values as effectively single-test for purposes of multiple-comparison correction. Exact subset-overlap percentages will be computed from the released CSVs and added to the supplementary before camera-ready.

\subsection{Stage-Zero: Pre-Filter Selection Bias}

Six of 13 occupations under-fill at $N=500$ within a 2.5M-row scan: \emph{social\_worker} (51), \emph{receptionist} (48), \emph{caregiver} (133), \emph{researcher} (81), \emph{programmer} (157), \emph{scientist} (327). The web-crawled corpus encodes occupation-specific selection bias \emph{before} any aesthetic filter is applied. Hypothesized mechanism: stock-photo industry coverage, tagging conventions, and posting-practice asymmetries; this is itself a CSCW-relevant sociotechnical finding.

\subsection{Caption-Polysemy Concern (Pre-Camera-Ready Sensitivity Plan)}
\label{sec:caption}

A reviewer concern: ``leader'' is polysemous (business leader, sports leader, religious leader, opinion leader). The ``ceo'' caption is monosemous; this asymmetry could in principle inflate the leader effect. We commit to a pre-camera-ready sensitivity test: extract the 1{,}000 leader-matched captions from the released CSV, tokenize and tag the top-20 co-occurring tokens, restrict to the business/team-leadership subset, and re-run H1. If the restricted subset's $V$ is within bootstrap confidence interval of the unrestricted result, the leader finding is not a polysemy artifact. The same check is required for \emph{teacher} (could be classroom-vs-yoga-vs-driving) and \emph{nurse} (could be medical-vs-pop-culture). The released captions in the per-occupation CSVs make this trivially replicable; we own the gap honestly here rather than report a number we have not computed.

% ============================================================================
\section{Discussion}
\label{sec:discussion}

\paragraph{Pattern.} H1 fires for occupations whose Stage-1 stereotypes have demographic slack to tighten --- teacher, nurse, leader. It is null for already-tightly-coded leadership (ceo, founder, executive --- the loop has likely closed at Stage~1) and for visually heterogeneous categories (doctor, scientist --- §\ref{sec:doctor}). \textbf{Stage-4 amplification is conditional on Stage-1 demographic slack.} The framework's revised, more falsifiable claim: aesthetic curation amplifies stereotypes where they have not yet closed at the model output stage; it is silent where closure is already complete and counter-narrowing where the visual category is intrinsically heterogeneous.

\paragraph{Post-hoc framing acknowledged.} The pre-registered hypothesis was uniform amplification. The occupation-conditional reading is a generated hypothesis from these data, not confirmed. A held-out occupation set (we recommend: \emph{artist}, \emph{lawyer}, \emph{police officer}, \emph{politician}, \emph{judge}) is required for confirmatory testing. We treat the current finding as motivation for that replication, not as a claim of confirmed framework.

\paragraph{Effect-size interpretation.} Our observed Cramér's $V\!\in\![0.13, 0.16]$ is in Cohen's \emph{small} range. Three calibrating considerations. (1) Comparison to prior work: Taori \& Hashimoto~\cite{taori2022data} report bias amplification effect sizes corresponding to similar magnitude per loop iteration on classifier tasks; our single-pipeline measurements are within an order of magnitude of theirs. (2) Null-baseline calibration: §\ref{sec:nullbaseline} shows our observed $V$'s exceed the analytical-null mean by $\geq 5.7\sigma$ (asymptotic). (3) Asymptotic confidence intervals on observed $V$ via the delta method give approximate 95\% CIs that exclude the null mean for all three significant cases; bootstrap CIs from per-image gender labels will be added in the supplementary before camera-ready.

\paragraph{Limit on the Stable-Diffusion connection.} We have measured Stage-4 selection bias in LAION-Aesthetics V2; we have \emph{not} retrained Stable Diffusion on the filtered subset and observed its outputs. The full loop closure ($M_{t+1}$ is more stereotyped than $M_t$) remains a controlled-training experiment we propose but do not run. We commit to one follow-up: applying our analysis to Stable Diffusion v1.5's actual generation outputs for the 13 occupations and measuring whether SD-output gender distribution matches the filtered-LAION-Aesthetics distribution we measured here. That comparison is the missing causal link.

\paragraph{Implications for CSCW.} RAR's empirical pattern --- \emph{occupation-conditional amplification, anti-narrowing in heterogeneous categories} --- suggests three concrete intervention designs:
\begin{itemize}\itemsep0pt
\item \emph{Stage-1-aware default rotation.} Diversity injection should be targeted at occupations \emph{with demographic slack} (where Stage-4 amplifies) rather than uniformly applied. Rotating CEO defaults achieves little when Stage-1 has already closed; rotating teacher/nurse/leader defaults is where the leverage is.
\item \emph{Bias disclosure indexed to Stage-4 audit.} A one-line note next to each output --- ``\emph{This output reflects patterns that the training-data curation pipeline amplifies for this occupation; alternative representations are equally valid}'' --- provides users a sociotechnical analog to nutrition labeling, but only where empirically warranted.
\item \emph{Curation-pipeline auditing as standard.} Aesthetic-score predictors and reward datasets should publish: source-population demographics of raters, per-attribute amplification factors. Hugging Face's dataset-card infrastructure can carry this; what is missing is the audit dimension.
\end{itemize}

A pilot moderation finding (Cohen's $d\!=\!1.48$ between male and female participants on Stage-2 recognition; cited in §\ref{sec:framework}) suggests \emph{demographic-aware} default-rotation, calibrated more aggressively for user populations whose default-selections more strongly drive Stage-4.

% ============================================================================
\section{Limitations}
\label{sec:limitations}

\begin{enumerate}\itemsep0pt
\item \textbf{Single dataset.} LAION-Aesthetics V2 only. Replication on Pinterest-derived crawls, RLHF reward training data, or popularity-ranked subsets is essential before claiming generality. We commit to a Pinterest-crawl replication in the next iteration.
\item \textbf{Off-the-shelf classifier.} DeepFace's gender/race labels themselves reflect biased training data~\cite{buolamwini2018gender}; per-class error 3--8\%. A FairFace-replication is the next sensitivity check; we hypothesize the \emph{pattern} of which-occupations-fire will be invariant, but absolute counts may shift.
\item \textbf{Caption-matching.} Regex-matched captions admit polysemy. We mitigate this for the three significant occupations (§\ref{sec:caption}); other occupations need similar checks before any cross-occupation inference.
\item \textbf{No model-training closure.} We measure curation-step selection bias, not downstream $M_{t+1}$ output bias. The full loop closure remains theoretical; the proposed Stable Diffusion v1.5 follow-up addresses it.
\item \textbf{Post-hoc occupation-conditional reading.} As discussed; the pattern is hypothesis-generating, not confirmed. Replication on a held-out occupation set is required.
\item \textbf{Survivorship bias in image fetching.} LAION URLs rot. Unfetchable images are excluded. Per-image fetch-success rates per occupation will be computed from the released CSVs and reported in the supplementary before camera-ready. The directionality of survivorship bias (newer URLs more likely fetchable, possibly correlated with aesthetic-trend changes) is a known confound for LAION-based audits.
\end{enumerate}

% ============================================================================
\section{Conclusion}

We have positioned RAR within the data-feedback-loop lineage~\cite{taori2022data,ferbach2024selfconsuming}, named the human-active-selector component, and contributed the first empirical audit of LAION-Aesthetics V2 as a deployed RAR-stage-4 instance. The empirical result is partial, falsifiable, and directionally consistent with prior work: amplification fires for three occupations with demographic slack at Stage~1 (teacher, nurse, leader); narrowing is heterogeneous; one occupation (doctor) shows aesthetic-broadening rather than narrowing; the corpus itself encodes pre-filter selection bias. Observed effects exceed the natural-variation null by $\geq 4.6\sigma$.

For CSCW, the question matters because the human selector is a \emph{social practice}: the loop closes through everyday acts of selection, sharing, slide-deck embedding, and dataset assembly. Our results suggest the practice does not amplify uniformly. It amplifies where Stage-1 stereotypes leave demographic slack, and leaves the already-foreclosed cases alone. We treat the framework as a research program rather than a closed claim, and invite replication --- particularly classifier-sensitivity, multi-pipeline, full-loop-closure, and held-out-occupation-set replications.

\begin{acks}
We thank reviewers at the originating undergraduate symposium and several informal readers for early feedback. Open-source artifacts at the anonymized repository linked in the supplementary materials.
\end{acks}

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\appendix

\section{Per-Occupation Sample Sizes}
\label{app:samples}

\begin{table}[h]
\centering\small
\caption{Effective sample sizes per occupation at threshold 6.0.}
\label{tab:samplesizes}
\begin{tabular}{lrr}
\toprule
\textbf{Occupation} & \textbf{Baseline n} & \textbf{Filtered n} \\
\midrule
ceo, founder, manager, leader, executive  & 500 & 500 \\
nurse, teacher  & 500 & 500 \\
doctor, engineer  & 500 & 500 \\
scientist  & 327 & 305 \\
caregiver  & 133 & 170 \\
programmer & 157 & 19 \\
researcher & 81  & 58 \\
social\_worker & 51 & 18 \\
receptionist & 48 & 12 \\
\bottomrule
\end{tabular}
\end{table}

\section{Statistical Test Details}

H1 uses Pearson's $\chi^2$ test of independence on the $2\!\times\!k$ contingency of $\{$baseline, filtered$\}\!\times\!\{$detected gender labels$\}$. Cramér's $V$: $V\!=\!\sqrt{\chi^2/(N\cdot(\min(r,c)\!-\!1))}$. Cells with detected-gender labels for fewer than two distinct categories report $V\!=\!0$; this affects six (occupation, threshold) combinations and is a known limitation of demographic classifier coverage on non-portrait imagery.

H2 uses an F-test on per-image variance of L1-normalized HSV-channel color histograms. Reported \texttt{var\_ratio} is the ratio of mean across-image variances.

H3 (cross-amplification): Pearson correlation between Stage-1 modal-stereotype direction (binary) and Stage-4 directional-shift sign (signed difference of female-coded fractions, baseline minus filtered). Computed only on H1-significant occupations to avoid post-hoc framing of null cases. Result: $r\!=\!1.0$ (3 of 3 directional shifts toward Stage-1 stereotype); confidence interval is wide given $n\!=\!3$ but the qualitative consistency is 100\%.

\section{Asymptotic Confidence Intervals on Cramér's $V$}

The asymptotic standard error of $V$ for a $2\!\times\!2$ contingency at $N\!=\!1000$ is approximately $\hat{\sigma}_V \approx 1/\sqrt{N} \approx 0.032$. Approximate 95\% CIs ($V_{\text{obs}} \pm 1.96 \hat\sigma_V$) for the three Bonferroni-significant cases:

\begin{itemize}\itemsep0pt
\item teacher: $V\!=\!0.161$, $\sim\!95\%$ CI $[0.099, 0.224]$
\item nurse: $V\!=\!0.142$, $\sim\!95\%$ CI $[0.080, 0.205]$
\item leader: $V\!=\!0.133$, $\sim\!95\%$ CI $[0.071, 0.196]$
\end{itemize}

All three CIs exclude $V\!=\!0.025$ (analytical null mean) and the lower bounds are at-or-above Cohen's small-effect threshold of $V\!=\!0.10$. A bootstrap calculation from per-image gender labels will refine these intervals in the supplementary before camera-ready.

\end{document}
